---
title: "tas_cmip6_nasa_giss"
author: "Kami Makabe"
date: "4/15/2021"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### load .nc files
```{r}
library(raster)
library(ncdf4)
tas126_1<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp126_r1i1p1f2_gn_201501-205012.nc")
tas126_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp126_r1i1p1f2_gn_205101-210012.nc")
tas245<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp245_r1i1p1f2_gn_201501-205012.nc")
tas245_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp245_r1i1p1f2_gn_205101-210012.nc")
tas585_1<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp585_r1i1p1f2_gn_201501-205012.nc")
tas585_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/tas_Amon_GISS-E2-1-G_ssp585_r1i1p1f2_gn_205101-210012.nc")

```

### get geom boundary from natural earth
```{r}
library(rnaturalearth)
Malawi<-ne_countries(country = 'malawi', returnclass = 'sf')
```


### subset
```{r}
library(raster)
tas126Malawi<-crop(tas126_1, Malawi)
tas126Malawim<-mask(tas126Malawi, Malawi)
tas126Malawi2<-crop(tas126_2, Malawi)
tas126Malawim2<-mask(tas126Malawi2, Malawi)

tas245Malawi<-crop(tas245, Malawi)
tas245Malawim<-mask(tas245Malawi, Malawi)
tas245Malawi2<-crop(tas245_2, Malawi)
tas245Malawim2<-mask(tas245Malawi2, Malawi)

tas585Malawi<-crop(tas585_1, Malawi)
tas585Malawim<-mask(tas585Malawi, Malawi)
tas585Malawi2<-crop(tas585_2, Malawi)
tas585Malawim2<-mask(tas585Malawi2, Malawi)
```

### convert to data frame

```{r, warning=FALSE, message=FALSE}
library(raster)
library(reshape)
library(base)
library(dplyr)
library(tidyr)
dftas126<-as.data.frame(tas126Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_tas126<-as.data.frame(tas126Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))

dftas245<-as.data.frame(tas245Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_tas245<-as.data.frame(tas245Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))

dftas585<-as.data.frame(tas585Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_tas585<-as.data.frame(tas585Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
```

### combine dataframes for each scenario type and remove null/missing values
```{r}
df_tas126comb<-rbind(dftas126,df2_tas126)
df_tas126comb<-na.omit(df_tas126comb)

df_tas245comb<-rbind(dftas245,df2_tas245)
df_tas245comb<-na.omit(df_tas245comb)

df_tas585comb<-rbind(dftas585,df2_tas585)
df_tas585comb<-na.omit(df_tas585comb)
```

### add column 'Scenario' to dataframes
```{r}
Scenario<-'ssp126'
tas1<-cbind(df_tas126comb,Scenario)
Scenario<-'ssp245'
tas2<-cbind(df_tas245comb,Scenario)
Scenario<-'ssp585'
tas5<-cbind(df_tas585comb,Scenario)
tascomb<-rbind(tas1,tas2,tas5)
```

## extract and format the date values into as.date

```{r, include=FALSE}
library(stringr)
Dt<-tascomb$variable
str_sub(Dt,-2)
str_sub(Dt,-5,-4)
str_sub(Dt,-10,-7)
paste(str_sub(Dt,-10,-7),str_sub(Dt,-5,-4),str_sub(Dt,-2), sep = "-")
Date<-as.Date(paste(str_sub(Dt,-10,-7),str_sub(Dt,-5,-4),str_sub(Dt,-2), sep = "-"))

```

#### add proper date column to your dataframe

```{r}
library(lubridate)
tascombdt<-cbind(tascomb,Date)
```
### add Month and year columns

```{r}
Year<-substr(Dt,2,5)
Month<-substr(Dt,7,8)
Day<-substr(Dt,10,11)
tasdf_nasa<-cbind(tascombdt, Year, Month,Day)
```

### rename column 'value' to 'tas'

```{r}
colnames(tasdf_nasa)[colnames(tasdf_nasa)=="value"]<-"tas"
```

### Plot monthly trend
```{r}

library(ggplot2)
ggplot(tasdf_nasa, aes(Date,tas))+
  geom_line(col='black')+
  facet_grid(Scenario~.,scales = 'free_x')+
  #scale_alpha_manual(breaks = 50)+
  ggtitle('Monthly Temperature Trend')+
  theme_bw()
```

### annual mean time series

```{r}
library(dplyr)
tasanndat<-tasdf_nasa%>% group_by(Scenario, Year)%>% 
  summarise(across(contains("tas"), ~mean(tas, na.rm=TRUE)))
```
  
#plot annual tas trend

```{r}
library(ggeasy)
qplot(Year, tas, data = tasanndat, geom = "line", group = Scenario) +
     facet_grid(Scenario ~ ., scale = "free_x") +
     scale_x_discrete() +
  ggtitle('Annual Temperature Trend')+
     easy_rotate_x_labels(angle = 90, side = "right")
```