---
title: "pr_cmip6_nasa_giss"
author: "Kami Makabe"
date: "4/15/2021"
output: html_document
---


```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### load .nc file
```{r}
library(raster)
library(ncdf4)
ssp126_1<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp126_r1i1p1f2_gn_201501-205012.nc")
ssp126_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp126_r1i1p1f2_gn_205101-210012.nc")
ssp245<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp245_r1i1p1f2_gn_201501-205012.nc")
ssp245_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp245_r1i1p1f2_gn_205101-210012.nc")
ssp585_1<-stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp585_r1i1p1f2_gn_201501-205012.nc")
ssp585_2<- stack("C:/Workspace/RMardown_NAPS/Open-NAPs-Data/CMIP6/pr_Amon_GISS-E2-1-G_ssp585_r1i1p1f2_gn_205101-210012.nc")

```

### get geom boundary from natural earth
```{r}
library(rnaturalearth)
Malawi<-ne_countries(country = 'malawi', returnclass = 'sf')
```


### subset
```{r}
library(raster)
ssp126Malawi<-crop(ssp126_1, Malawi)
ssp126Malawim<-mask(ssp126Malawi, Malawi)
ssp126Malawi2<-crop(ssp126_2, Malawi)
ssp126Malawim2<-mask(ssp126Malawi2, Malawi)

ssp245Malawi<-crop(ssp245, Malawi)
ssp245Malawim<-mask(ssp245Malawi, Malawi)
ssp245Malawi2<-crop(ssp245_2, Malawi)
ssp245Malawim2<-mask(ssp245Malawi2, Malawi)

ssp585Malawi<-crop(ssp585_1, Malawi)
ssp585Malawim<-mask(ssp585Malawi, Malawi)
ssp585Malawi2<-crop(ssp585_2, Malawi)
ssp585Malawim2<-mask(ssp585Malawi2, Malawi)
```

### convert to data frame

```{r}
library(raster)
library(reshape)
library(base)
library(dplyr)
library(tidyr)
dfssp126<-as.data.frame(ssp126Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_ssp126<-as.data.frame(ssp126Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))

dfssp245<-as.data.frame(ssp245Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_ssp245<-as.data.frame(ssp245Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))

dfssp585<-as.data.frame(ssp585Malawim, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
df2_ssp585<-as.data.frame(ssp585Malawim2, xy=TRUE)%>%
  melt(id.vars=c('x','y'))
```

### combine dataframes for each scenario type and remove null/missing values
```{r}
df_ssp126comb<-rbind(dfssp126,df2_ssp126)
df_ssp126comb<-na.omit(df_ssp126comb)

df_ssp245comb<-rbind(dfssp245,df2_ssp245)
df_ssp245comb<-na.omit(df_ssp245comb)

df_ssp585comb<-rbind(dfssp585,df2_ssp585)
df_ssp585comb<-na.omit(df_ssp585comb)
```

### add column 'Scenario' to dataframes
```{r}
Scenario<-'ssp126'
xs1<-cbind(df_ssp126comb,Scenario)
Scenario<-'ssp245'
xs2<-cbind(df_ssp245comb,Scenario)
Scenario<-'ssp585'
xs5<-cbind(df_ssp585comb,Scenario)
xscomb<-rbind(xs1,xs2,xs5)
```

## extract and format the date values into as.date

```{r, include=FALSE}
library(stringr)
Dt<-xscomb$variable
str_sub(Dt,-2)
str_sub(Dt,-5,-4)
str_sub(Dt,-10,-7)
paste(str_sub(Dt,-10,-7),str_sub(Dt,-5,-4),str_sub(Dt,-2), sep = "-")
Date<-as.Date(paste(str_sub(Dt,-10,-7),str_sub(Dt,-5,-4),str_sub(Dt,-2), sep = "-"))

```

#### add proper date column to your dataframe

```{r}
library(lubridate)
xscombdt<-cbind(xscomb,Date)
```
### add Month and year columns

```{r, include=FALSE}
Year<-substr(Dt,2,5)
Month<-substr(Dt,7,8)
Day<-substr(Dt,10,11)
prdf_nasa<-cbind(xscombdt, Year, Month,Day)
```

### rename column 'value' to 'pr'

```{r}
colnames(prdf_nasa)[colnames(prdf_nasa)=="value"]<-"pr"
```

### Plot monthly trend
```{r}

library(ggplot2)
ggplot(prdf_nasa, aes(Date,pr))+
  geom_line(col='black')+
  facet_grid(Scenario~.,scales = 'free_x')+
  ggtitle('Monthly Precipitation')+
  theme_bw()
```

### annual mean time series

```{r}
library(dplyr)
pranndat<-prdf_nasa%>% group_by(Scenario, Year)%>% 
  summarise(across(contains("pr"), ~mean(pr, na.rm=TRUE)))
```
  
#plot Annual pr 
```{r}
library(ggeasy)
qplot(Year, pr, data = pranndat, geom = "line", group = Scenario) +
     facet_grid(Scenario ~ ., scale = "free_x") +
     scale_x_discrete() +
  ggtitle('Annual Precipitation Trend')+
     easy_rotate_x_labels(angle = 90, side = "right")
```